Solution of hydrodynamic equilibrium in black-oil reservoirs: the classical material balance equation

### 2015-04-07 pituba/article/article_spe.tex
fixing+

\begin{abstract}
This work examines underlying details in obtaining the classical material balance equation from a three dimensional black-oil model typical of reservoir simulations. The black-oil equations are integrated in reservoir volume and quantities are spatially averaged. The material balance equation is obtained for averaged quantities and it is applied to calculate reservoir pressure. It is shown that the calculated pressure corresponds to the local pressure at the datum level. The definition of a convenient datum level is also discussed. It is found that the physical correctness of PVT properties
% obtained from laboratory experiments or empirical correlations
is a sufficient condition for solution uniqueness of the material balance equation when pressure is the unknown quantity.
Finally, a numerical scheme for solving the material balance equation is proposed based on nonlinear root finding algorithms.
\end{abstract}

@@

We assume that the reservoir is a hydraulically connected volume where hydrocarbon saturation is nonzero. By a hidraulically connected volume, we mean that it is possible to observe a common pressure decline trend when any reservoir portion is under production \citep{dakepractice}. Reservoir fluids are composed of oil and gas pseudo-components and water component. There can be three fluid phases at most: oil, gas and (liquid) water. Oil phase is composed of oil and gas components, gas phase is composed uniquely by gas component and, water phase, uniquely by water component. The only allowed mass transfer is of gas component between oil and gas phases. The gas phase composition is always that of a dry gas.

@@

It follows from Duhem's theorem that the equilibrium of oil and gas phases is determined if, in addition to the mass balance equations, two other properties are prescribed. From Gibbs' phase rule, the system has two degrees of freedom. It is then necessary to specify two intensive properties: the constant temperature is one and the local pressure is chosen as the other. Since temperature was assumed constant, oil phase density becomes a function of pressure alone. The same is valid for mass fractions of gas and oil components in the oil phase:
\begin{equation}
\rho_o = \rho_o \left(p_o\right) \, , \quad Y_g = Y_g (p_o) \, , \quad Y_o = Y_o (p_o) \, .
\end{equation}

When the oil phase is undersaturated, the system has an additional degree of freedom and both pressure and the amount of dissolved gas must be known to define its thermodynamic state, e.g., $\rho = \rho(p_o,Y_o)$.

@@

Eqs. \eqref{eq: Sw2}, \eqref{eq: So2} and \eqref{eq: Sg2} are rewritten in terms of averaged quantities:

@@

We now introduce usual definitions from petroleum engineering for the description of volumetric and solubility changes. In what concerns volumetric changes, $B_w$, $B_o$ and $B_g$, the formation volume factors of water, oil and gas phases, respectively, express the ratio of a phase volume at reservoir condition to its volume at standard surface condition, i.e.:

@@

In what concerns changes in gas solubility in oil phase, the solubility ratio ($R_s$) gives the ratio between the volume of dissolved gas and the volume of oil both measured at surface condition:

@@

An equation for the bubble-point can be obtained by setting $\bar{S}_{g2}=0$ in Eq.~\eqref{eq: Sg5} and solving it for the solubility ratio at bubble-point pressure, $R_s(p=P_b)$:
\begin{equation}\label{eq: eqPb}
\begin{split}
R_{s2}\left(p=P_b\right) = &\Big[N\left(m B_{1}/B_{g1} +R_{s1}\right)\\
&+G_i-G_p\Big]/\left(N-N_p\right) \, .
\end{split}
\end{equation}

@@

Eq~\eqref{eq: MBE} involves fluid properties, original phase volumes ($N$, $W$ and $G)$, cumulative production and injection and aquifer influx.
It is an algebraic equation that can be solved for any one of its terms, provided that all remaining are known. This works focuses on its solution for pressure.

@@

@@

\begin{equation}
\rho_o\left( t,z\right) \approx \rho_o\left(t,z_{datum}\right)\left[1+\frac{1}{\rho_o}\frac{\partial \rho_o}{\partial z} dz\right]
\end{equation}
For an undersaturated oil with no compositional gradient,
 %As pressure increases with depth according to the hydrostatic gradient, density also increases but by a negligible amount. This approximation can be quantified by putting together the expressions for compressibility and hydrostatic equilibrium. For an undersaturated oil with no compositional gradient,
 \begin{equation}
\frac{1}{\rho_o}\left(\frac{\partial \rho_o}{\partial z} \right)_{T,Y_o} = \frac{1}{\rho_o}\left(\frac{\partial \rho_o}{\partial p_o} \right)\frac{\partial p_o}{\partial z} = c_o \rho_o g \, ,
 \end{equation}
 where $c_o$ is the isothermal compressibility.

 For a saturated oil,
 \begin{equation}
\frac{1}{\rho_o}\left(\frac{\partial \rho_o}{\partial z} \right)_T = \frac{1}{\rho_o}\left(\frac{\partial \rho_o}{\partial p_o} \right)_T = 
\frac{1}{\rho_o}\frac{d}{d p_o}\left(\frac{\rho_o^{ST}+R_s\rho_g^{ST}}{B_o}\right) 
 \end{equation}

@@

From Eq.~\eqref{eq: dRsdz2}, we see that the variation of pressure with depth may not be negligible for the phase equilibrium,  specially for thick reservoirs of light oil. In such cases, using an averaged value for the solubility ratio may be worth the accuracy gain in computing dissolved gas volume.

>>>>

\begin{abstract}
It is generally known that the material balance equation can be derived from a 3D black-oil model typical of reservoir simulation. However, the existing literature is not comprehensive about what are the needed assumptions and how the pressure calculated from the material balance equation relates to the pressure field of a 3D black-oil model.

This work examines underlying details in obtaining the material balance equation from a 3D black-oil model. The black-oil equations are integrated in reservoir volume and quantities are averaged in space. The material balance equation is obtained for the averaged quantities and it is applied to calculate reservoir pressure. It is shown that this calculated pressure corresponds to the pressure at a chosen datum level in the 3D black-oil model (and not, in general, to the average pressure in the 3D model). The definition of a convenient datum level is also discussed. It is found that the physical correctness of PVT properties 
is a sufficient condition for solution uniqueness of the material balance equation when pressure is the unknown quantity.
Finally, a numerical scheme for solving the material balance equation is proposed based on nonlinear root finding algorithms.
\end{abstract}

@@

We assume that the reservoir is a hydraulically connected volume where hydrocarbon saturation is nonzero. By a hidraulically connected volume, we mean that it is possible to observe a common pressure decline trend when any reservoir portion is under production \citep{dakepractice}. Reservoir fluids are composed of oil and gas pseudo-components and water component. There can be three fluid phases at most: oil, gas and liquid water. Oil phase is composed of oil and gas components, gas phase is composed uniquely by gas component and, water phase, uniquely by water component. The only allowed mass transfer is of gas component between oil and gas phases. The gas phase composition is always that of a dry gas.

@@

It follows from Duhem's theorem that the equilibrium of oil and gas phases is determined if, in addition to the mass balance equations, two other properties are prescribed. From Gibbs' phase rule, the oil an gas system has two degrees of freedom. It is then necessary to specify two intensive properties. The constant temperature is one and the local pressure is chosen as the other. Since temperature was assumed constant, oil phase density becomes a function of pressure alone. The same is valid for mass fractions of gas and oil components in the oil phase:
\begin{equation}
\rho_o = \rho_o \left(p_o\right) \, , \quad Y_g = Y_g (p_o) \, , \quad Y_o = Y_o (p_o) \, .
\end{equation}

When the oil phase is undersaturated, the system has one additional degree of freedom and both pressure and the amount of dissolved gas must be known to define its thermodynamic state, e.g., $\rho_o = \rho_o (p_o,Y_o)$.

@@

Eqs. \eqref{eq: Sw2}, \eqref{eq: So2} and \eqref{eq: Sg2} can now be rewritten in terms of averaged quantities:

@@

Usual definitions from petroleum engineering can be introduced for the description of volumetric and solubility changes. For volumetric changes, $B_w$, $B_o$ and $B_g$, the formation volume factors of water, oil and gas phases, respectively, express the ratio of a phase volume at reservoir condition to its volume at standard surface condition, i.e.:

@@

For changes in gas solubility in oil phase, the solubility ratio ($R_s$) gives the ratio between the volume of dissolved gas and the volume of oil both measured at surface condition:

@@

An inplicit equation for the bubble-point can be obtained by setting $\bar{S}_{g2}=0$ in Eq.~\eqref{eq: Sg5} and solving it for the solubility ratio at bubble-point pressure, $R_s(p=P_b)$:
\begin{equation}\label{eq: eqPb}
\begin{split}
R_{s2}\left(p=P_b\right) = &\Big[N\left(m B_{1}/B_{g1} +R_{s1}\right)\\
&+G_i-G_p\Big]/\left(N-N_p\right) \, .
\end{split}
\end{equation}

@@

Eq~\eqref{eq: MBE} involves fluid properties, original phase volumes ($N$, $W$ and $G)$, cumulative production and injection and aquifer influx.
It is an algebraic equation that can be solved for any one of its terms, provided that all remaining are known. This work focuses on its solution for pressure.

@@

\begin{itemize}
\item density is approximately constant;
\item pressure field does not deviates from hydrostatic equilibrium.
\end{itemize}

@@

\begin{equation}
\rho_o\left( t,z\right) \approx \rho_o\left(t,z_{datum}\right)\left[1+\frac{1}{\rho_o}\frac{\partial \rho_o}{\partial z} dz\right]
\end{equation}
For an undersaturated oil, $\rho_o$ is a function of pressure and the amount of dissolved gas:
 %As pressure increases with depth according to the hydrostatic gradient, density also increases but by a negligible amount. This approximation can be quantified by putting together the expressions for compressibility and hydrostatic equilibrium. For an undersaturated oil with no compositional gradient,
 \begin{equation}
\frac{1}{\rho_o}\left(\frac{\partial \rho_o}{\partial z} \right)_{T} = \frac{1}{\rho_o}\left(\frac{\partial \rho_o}{\partial p_o} \frac{\partial p_o}{\partial z} + \frac{\partial \rho_o}{\partial Y_o}\frac{\partial Y_o}{\partial z}\right) \, . 
 \end{equation}
Neglecting any compositional gradient, $\partial Y_o/\partial z = 0$, and using the hydrostatic gradient:
\begin{equation}
\frac{1}{\rho_o}\left(\frac{\partial \rho_o}{\partial z} \right)_{T} = c_o \rho_o g \, ,
\end{equation}
 where $c_o$ is the isothermal compressibility.

 For a saturated oil,
 \begin{equation}
\frac{1}{\rho_o}\left(\frac{\partial \rho_o}{\partial z} \right)_T = g\left(\frac{\partial \rho_o}{\partial p_o} \right)_T = 
g\frac{d}{d p_o}\left(\frac{\rho_o^{ST}+R_s\rho_g^{ST}}{B_o}\right) 
 \end{equation}

@@

From Eq.~\eqref{eq: dRsdz2}, we see that the variation of pressure with depth may not be negligible for the phase equilibrium,  specially for thick reservoirs of low saturation pressure. In such cases, using an averaged value for the solubility ratio may be worth the accuracy gain in computing dissolved gas volume.
